Statistical Modeling of Galaxy Catalogues with Normalizing Flows

### 2022-12-01 src/tex/ms.tex
PhotErr.

@@

\begin{abstract}
    Normalizing flows are powerful tools for learning high-dimensional probability distributions from samples thereof.
    We introduce PZFlow, a Python package for the statistical modeling of tabular data using normalizing flows.
    We use PZFlow to model the seven dimensional space of galaxy redshift and LSST photometry.
    Using these normalizing flows, we forward model galaxy catalogs in which each galaxy property has an associated ``true'' posterior distribution.
    We discuss the importance of these posteriors for the comprehensive evaluation of the redshift posteriors produced by photometric redshift (photo-z) estimators.
    We also demonstrate the use of an ensemble of normalizing flows for photo-z estimation.
    While we focus on photo-z estimation and validation, we emphasize that these methods are applicable to any galaxy properties, or any other tabular data.
\end{abstract}

@@

\begin{figure*}[t!]
    \script{intro/plot_two_moons.py}
    \begin{centering}
        \includegraphics{figures/twomoons_example.pdf}
        \caption{
            A normalizing flow demonstrated on the two moons data set from \texttt{scikit-learn}.
            The two moons data on the left is mapped onto a two dimensional uniform distribution by the bijection $f$.
            The data are colored by quadrant to visualize their image in the latent space.
            You can sample the data distribution by sampling from the uniform distribution, and using $f^{-1}$ to map the samples back to the data space.
        }
        \label{fig:two-moons}
    \end{centering}
\end{figure*}

@@

Figure \ref{fig:two-moons} shows an example of a normalizing flow that transforms the \texttt{scikit-learn} two moons distribution into a uniform distribution.
The data points are colored by quadrant to visualize their image under $f$.

@@

PZFlow is a package for building normalizing flows in Python, with a focus on high-performance out-of-the-box modeling of high dimensional tabular data.
Data is handled in Pandas DataFrames \citep{pandas} while the normalizing flows are implemented in Jax \citep{jax}, which allows efficient, parallelizable, GPU-enabled calculations for very large data sets.
The code is easily installable from the Python Package Index\footnote{\url{https://pypi.org/project/pzflow/}} (PyPI) and is hosted on Github,\footnote{\url{https://github.com/jfcrenshaw/pzflow}} including several tutorial notebooks.

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\begin{figure}[t!]
    \script{photo-z/plot_ensemble_losses.py}
    \begin{centering}
        \includegraphics{figures/ensemble_losses.pdf}
        \caption{
            Training losses for the four flows in the flow ensemble.
            You can see that each of the flow achieves a different minimum loss.
            After epochs 50 and 100, you can see a drop in the loss due to the decrease in the learning rate.
        }
        \label{fig:ensemble-losses}
    \end{centering}
\end{figure}

@@

We estimate photometric errors for LSST using the error model from \citet{ivezic2019}, except we do not make the high SNR approximation used there.

\citet{ivezic2019} calculates a Gaussian photometric error in magnitude space using Equation 5 from that paper: 
\begin{align}
    \sigma_\text{rand}^2 = (0.04 - \gamma) x + \gamma x^2 ~~ (\text{mag}^2),
\end{align}
where $x \equiv 10 ^ {0.4(m - m_5)}$, $m_5$ is the 5$\sigma$ depth for the band, and $\gamma$ is a band-dependent parameter listed in Table 2 of the same paper.
This formula is based on the approximation
\begin{align}
    \sigma_\text{rand} = 2.5 \log_{10} \left(1 + \frac{N}{S} \right) \approx \frac{N}{S},
\end{align}
where $S$ and $N$ are the flux signal and noise, respectively.
This approximation is valid in the high signal-to-noise regime.

To generalize this model to the low signal-to-noise regime, we abandon this latter approximation.
For each magnitude, $m$, we calculate the corresponding flux, $S = 10^{-m/2.5}$, and the Gaussian flux noise,
\begin{align}
    N = S \, \sqrt{(0.04 - \gamma) x + \gamma x^2}, \label{eq:noise}.
\end{align}
For very bright galaxies, we impose a systematic floor corresponding to an error of 0.005 mags.
We then sample an observed flux, $S_\text{obs} \sim \mathcal{N}(S, N)$, and calculate the observed magnitude, $m_\text{obs} = -2.5 \log_{10} S_\text{obs}$.

For the reported photometric error, we do not use $N$, as this provides a deterministic link to the true magnitude via Equation \ref{eq:noise}.
Instead, we apply Equation \ref{eq:noise} again, this time using the observed magnitude instead of the true magnitude, yielding the estimated noise, $N_\text{obs}$.
Finally, we report the photometric error in magnitudes:
\begin{align}
    \sigma_m = 2.5 \log_{10} \left(1 + \frac{N_\text{obs}}{S_\text{obs}} \right).
\end{align}

>>>>

\author{Ziang Yan}

\author{Andrew Connolly}

@@

\begin{abstract}
    Normalizing flows are powerful tools for learning high-dimensional probability distributions from samples thereof, and have many applications in astronomy, including posterior estimation and forward modeling.
    We introduce PZFlow, a Python package for the statistical modeling of tabular data using normalizing flows.
    We use PZFlow to model photometric galaxy catalogs including redshifts, LSST photometry, size, and ellipticity.
    We generate synthetic galaxy catalogs in which each galaxy has a ``true'' redshift posterior distribution, and discuss the importance of these true posteriors for the comprehensive evaluation of the redshift posteriors produced by photometric redshift (photo-z) estimators.
    We also demonstrate the use of an ensemble of normalizing flows for photo-z estimation.
    While we focus on photo-z estimation and validation, we emphasize that these methods are applicable to any galaxy properties, and any other tabular data.
\end{abstract}

@@

\begin{figure*}[t!]
    \script{intro/plot_two_moons.py}
    \begin{centering}
        \includegraphics{figures/twomoons_example.pdf}
        \caption{
            A normalizing flow demonstrated on the two moons data set from scikit-learn.
            The two moons data on the left is mapped onto a two dimensional uniform distribution by the bijection $f$.
            The data are colored by quadrant to visualize their image in the latent space.
            You can sample the data distribution by sampling from the uniform distribution, and using $f^{-1}$ to map the samples back to the data space.
        }
        \label{fig:two-moons}
    \end{centering}
\end{figure*}

@@

Figure \ref{fig:two-moons} shows an example of a normalizing flow that transforms the scikit-learn \citep{sklearn} two moons distribution into a uniform distribution.
The data points are colored by quadrant to visualize their image under $f$.

@@

@@

\section{Conclusion \& Summary}
\label{sec:conclusion}

In this paper we introduced PZFlow, a normalizing flow package in Python, designed for the statistical modeling of astronomical data sets.
We used PZFlow to forward model galaxy catalogs that include photometry, redshifts, sizes, and ellipticities.
In addition, each galaxy in our catalog has a true redshift posterior, which can be convolved with measurement errors.
These true posteriors allow a direct evaluation of the posteriors produced by photo-z estimators.
A similar comparison can be made to posterior estimates for any other galaxy property.

Direct evaluation of photo-z posteriors is vital for future cosmology analyses which must use all of the information incorporated in the full redshift posteriors \citep{newman2022}.
This is because only the full redshift posteriors allow one to account for degeneracies in color-redshift space, which will otherwise bias cosmological inference.
Previous evaluations of photo-z performance focused on point estimates and metrics for ensembles of posteriors, but \citet{schmidt2020} showed these analyses are inadequate for modern cosmology.
This work enables a future analysis of photo-z estimators on a per-posterior basis, which is a major goal of the LSST DESC.

In addition to forward modeling galaxy catalogs, we demonstrated PZFlow's utility as a density estimator that can be applied to photo-z estimation and other statistical analyses.
PZFlow achieves high accuracy with relatively little fine tuning and with very few modeling assumptions.
PZFlow will be a powerful tool for the statistical analysis of tabular astronomical data.

Note that the code to reproduce the figures in this paper is hosted publicly on Github\footnote{\url{https://github.com/jfcrenshaw/pzflow-paper}}, and the code for each individual figure can be reached by clicking on the Github logo in the margin next to the figure.

\begin{acknowledgements}
    JFC is supported by...
\end{acknowledgements}

\software{
    adam \citep{adam},
    corner \citep{corner},
    dill \citep{dill},
    jax \citep{jax},
    jupyter \citep{jupyter},
    matplotlib \citep{matplotlib},
    numpy \citep{numpy},
    pandas \citep{pandas,pandas-software},
    scipy \citep{scipy},
    showyourwork \citep{showyourwork},
    scikit-learn \citep{sklearn}
}

@@

\begin{figure}[t!]
    \script{photo-z/plot_ensemble_losses.py}
    \begin{centering}
        \includegraphics{figures/ensemble_losses.pdf}
        \caption{
            Training losses for the four flows in the flow ensemble.
            We have zoomed in to the bottom of the loss curve so you can see that each of the flows converges to a different minimum loss.
        }
        \label{fig:ensemble-losses}
    \end{centering}
\end{figure}

@@

We estimate photometric errors for LSST using a generalization of the error model from \citet{ivezic2019}.
To derive the error model, we start with the noise-to-signal ratio (NSR) for an object with photon count $C$ and background noise $N_0$ (which depends on seeing, read-out noise, etc.):
\begin{align}
    \text{NSR}^2 = \frac{N_0^2 + C}{C^2}.
\end{align}
If we define $C=C_5$ when $\text{NSR}= 1/5$, then we can solve for $N_0$ and write
\begin{align}
    \text{NSR}^2 = \frac{1}{C_5} \left( \frac{C_5}{C} \right) + \left[ \left( \frac{1}{5} \right)^2 - \frac{1}{C_5} \right] \left( \frac{C_5}{C} \right)^2.
\end{align}
Defining $x = C_5/C = 10^{(m-m_5)/2.5}$ and $\gamma = 1/5^2 - 1/C_5$, we have
\begin{align}
    \text{NSR}^2 = (0.04 - \gamma) \, x + \gamma \, x^2 ~~ (\text{mag}^2),
\end{align}
which is Equation 5 from \citet{ivezic2019}.
Values for the band-dependent parameter $\gamma$ can be found in Table 2 of the same paper.

In the high signal-to-noise (SNR) limit, $\text{NSR} \ll 1$, and we can approximate
\begin{align}
    \sigma_\text{rand} = 2.5 \log_{10}(1 + \text{NSR}) \approx \text{NSR}.
    \label{eq:err}
\end{align}
This latter approximation is made by \citet{ivezic2019}, and errors are assumed to be Gaussian in magnitude space.
In contrast, we use the exact form of Equation \ref{eq:err}, and model errors as Gaussian in flux space.
Note that after the photometric errors are applied, the error is re-calculated from the ``observed'' flux, and this new error is reported as the estimated photometric error.
If the original photometric error were reported, it would provide a deterministic link to the original flux.

We have implemented this error model, along with several other extensions to it, in the Python package PhotErr, which is available on the Python Package Index\footnote{\url{https://pypi.org/project/photerr/}} (PyPI), and Github\footnote{\url{https://github.com/jfcrenshaw/photerr}}.
The extensions include different methods for handling non-detections, methods for modeling errors of extended objects \citep{vandenbusch2020,kuijken2019}, and error models for the Roman and Euclid space telescopes \citep{roman,euclid,graham2020}, however, these extensions are not used in this paper.
